---
title: "Compare Natura2000 with non-Natura2000 based on Angiosperm cube with dubicube"
author: "Ward Langeraert"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: right
    code-fold: true
    self-contained: true

editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(b3gbi) # v0.9.0
library(dubicube) # v0.12.3
library(glmmTMB)
library(emmeans)
library(cowplot)

# Globals
analysis_path <- here::here("_preprocessing/R/Natura2000")
data_path <- here::here("_preprocessing/data/raw")

cache_path <- file.path(analysis_path, "cache", "compare_natura2000_dubicube")
dir.create(cache_path, recursive = TRUE, showWarnings = FALSE)

out_path <- file.path(analysis_path, "output", "compare_natura2000_dubicube")
dir.create(out_path, recursive = TRUE, showWarnings = FALSE)
devices <- c("jpg", "tiff", "eps")

# Source
source(file.path(analysis_path, "source", "cache_compute.R"))
source(file.path(analysis_path, "source", "save_figure.R"))
```

# Goal

We test some things for Flanders use case of the B-Cubed project:
<https://flanders.b-cubed.eu/>

* Test uncertainty calculation dubicube and b3gbi on a simple indicator
* Calculate a new indicator and use effect classification to test significance
* Use area and sampling effort as correction factors

We focus on the "General indicators Natura 2000 angiospermae" ([link](https://flanders.b-cubed.eu/src/b3gbi_indicators_Natura2000.html)).
More specifically on the species richness within and outside Natura2000 habitat.

![Total number of unique species per year detected in Natura 2000 versus non Natura 2000 area visualised on the [dashboard](https://flanders.b-cubed.eu/src/b3gbi_indicators_Natura2000.html).](https://flanders.b-cubed.eu/src/b3gbi_indicators_Natura2000_files/figure-html/unnamed-chunk-2-1.png
)

# Data

We load the data and select the data from 1975 up to 2022.
We need to be able to calculate (1) species richness, (2) surface area, and (3) sampling effort.
We need (2) and (3) since we can assume that we get a higher species richness when a larger surface area is surveyed and when sampling effort is higher.

```{r}
data_natura2000 <- read_csv(
  file.path(data_path, "Angiosperm_data_flanders_Natura2000.csv"),
  name_repair = "minimal",
  col_select = -1,
  show_col_types = FALSE
) %>%
  filter(year %in% 1975:2022)

data_nonnatura2000 <- read_csv(
  file.path(data_path, "Angiosperm_data_flanders_nonNatura2000.csv"),
  name_repair = "minimal",
  col_select = -1,
  show_col_types = FALSE
) %>%
  filter(year %in% 1975:2022)

glimpse(data_natura2000)
```

1. We can calculate number of species

This figure is identical to the one on the dashboard (without the final years).

```{r}
sr_natura2000 <- data_natura2000 %>%
  summarise(n_spec = n_distinct(specieskey), .by = c("year")) %>%
  mutate(location = "Natura2000")
sr_nonnatura2000 <- data_nonnatura2000 %>%
  summarise(n_spec = n_distinct(specieskey), .by = c("year")) %>%
  mutate(location = "non-Natura2000")

bind_rows(sr_natura2000, sr_nonnatura2000) %>%
  ggplot(aes(x = year, y = n_spec, colour = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = "y ~ x", se = FALSE) +
  labs(x = "Year", y = "Number of species", colour = "Area type") +
  theme_minimal()
```

2. We can calculate surface area

We now each grid cell is 1 km².

```{r}
area_natura2000 <- data_natura2000 %>%
  summarise(area_km2 = n_distinct(eeacellcode), .by = c("year")) %>%
  mutate(location = "Natura2000")
area_nonnatura2000 <- data_nonnatura2000 %>%
  summarise(area_km2 = n_distinct(eeacellcode), .by = c("year")) %>%
  mutate(location = "non-Natura2000")

bind_rows(area_natura2000, area_nonnatura2000) %>%
  ggplot(aes(x = year, y = area_km2, fill = location)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Area (km²)", fill = "Area type") +
  theme_minimal()
```

3. We can almost calculate sampling effort

Normally we would calculate in the SQL query the count at a higher taxonomic level.
However, in this dataset the highest available is `ordercount`.
We could download a new cube that calculates count on a higher taxonomic level (perhaps across all taxa even) to get a better value.
For now, we just use the total number of occurrences per year over the whole datasets as a proxy for sampling effort.

The sum of the occurrences mixes biological abundance with observer effort.
If one habitat truly has more individuals, this metric may overestimate effort there.
If possible, a measure such as the number of visits or independent sampling events would be preferable.

```{r}
area_natura2000 <- data_natura2000 %>%
  summarise(total_occ = sum(occurrences), .by = c("year")) %>%
  mutate(location = "Natura2000")
area_nonnatura2000 <- data_nonnatura2000 %>%
  summarise(total_occ = sum(occurrences), .by = c("year")) %>%
  mutate(location = "non-Natura2000")

bind_rows(area_natura2000, area_nonnatura2000) %>%
  ggplot(aes(x = year, y = total_occ, colour = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = "y ~ x", se = FALSE) +
  labs(x = "Year", y = "Number of occurrences", colour = "Area type") +
  theme_minimal()
```

# Basic uncertainty calculation

Let's see how the **b3gbi** package handles uncertainty calculation.
We can add uncertainty using the `add_ci()` function (new from v0.9.0 onwards).
In the documentation, we read that confidence level calculation is not provided for observed species richness, since this is not relevant.

Natura2000:

```{r}
cube_natura2000 <- process_cube(data_natura2000)
obs_rich_natura2000 <- obs_richness_ts(cube_natura2000)
plot(obs_rich_natura2000)
```

non-Natura2000:

```{r}
cube_nonnatura2000 <- process_cube(data_nonnatura2000)
obs_rich_nonnatura2000 <- obs_richness_ts(cube_nonnatura2000)
plot(obs_rich_nonnatura2000)
```

Let's see what happens if we calculate confidence intervals with dubicube.

```{r}
my_spec_richness <- function(df) {
  df %>%
    summarise(diversity_val = n_distinct(taxonKey), .by = c("year"))
}
```

```{r}
# Bootstrapping (cached)
bootstrap_results_natura2000 <- cache_compute(
  path = cache_path,
  file = "bootstrap_results_natura2000",
  expr = {
    dubicube::bootstrap_cube(
      data_cube = cube_natura2000,
      fun = my_spec_richness,
      grouping_var = "year",
      samples = 1000,
      seed = 123
    )
  }
)
bootstrap_results_nonnatura2000 <- cache_compute(
  path = cache_path,
  file = "bootstrap_results_nonnatura2000",
  expr = {
    dubicube::bootstrap_cube(
      data_cube = cube_nonnatura2000,
      fun = my_spec_richness,
      grouping_var = "year",
      samples = 1000,
      seed = 123
    )
  }
)

# Interval calculation
ci_spec_richness_natura2000 <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = bootstrap_results_natura2000,
  grouping_var = "year",
  type = "perc",
  conf = 0.95
)
ci_spec_richness_nonnatura2000 <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = bootstrap_results_nonnatura2000,
  grouping_var = "year",
  type = "perc",
  conf = 0.95
)
```

As expected, the confidence intervals do not overlap the estimate (as explained [here](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-interval-calculation.html#issues-with-bias-correction-for-species-richness-indicators)).

```{r}
#| warning: false
# Convert bootstrap replicates to dataframe
bootstrap_results_natura2000_df <- boot_list_to_dataframe(
  boot_list = bootstrap_results_natura2000,
  grouping_var = "year"
) %>%
  mutate(location = "Natura2000",
         year = as.numeric(year))
bootstrap_results_nonnatura2000_df <- boot_list_to_dataframe(
  boot_list = bootstrap_results_nonnatura2000,
  grouping_var = "year"
) %>%
  mutate(location = "non-Natura2000",
         year = as.numeric(year))

# Get bias values
bias_df <- bind_rows(bootstrap_results_natura2000_df,
                     bootstrap_results_nonnatura2000_df) %>%
  distinct(year, estimate = est_original, `bootstrap estimate` = est_boot, location)

# Get estimate values
estimate_df <- bias_df %>%
  pivot_longer(cols = c("estimate", "bootstrap estimate"),
               names_to = "Legend", values_to = "value") %>%
  mutate(Legend = factor(Legend, levels = c("estimate", "bootstrap estimate"),
                         ordered = TRUE))
# Visualise
bind_rows(bootstrap_results_natura2000_df,
          bootstrap_results_nonnatura2000_df) %>%
  ggplot(aes(x = year)) +
  # Distribution
  geom_violin(aes(y = rep_boot, group = year),
              fill = alpha("cornflowerblue", 0.2)) +
  # Estimates and bias
  geom_point(data = estimate_df,
             aes(y = value, shape = Legend, colour = location),
             size = 1, alpha = 0.5) +
  # Intervals
  geom_errorbar(
    data = bind_rows(
      ci_spec_richness_natura2000 %>%
        mutate(location = "Natura2000",
               year = as.numeric(year)),
      ci_spec_richness_nonnatura2000 %>%
        mutate(location = "non-Natura2000",
               year = as.numeric(year))
    ),
    aes(ymin = ll, ymax = ul, colour = location),
    linewidth = 0.5
  ) +
  # Settings
  facet_wrap(~location, ncol = 1) +
  labs(y = "Species richness",
       x = "", shape = "Legend:", colour = "Location:") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Let's ignore the bias. We assume the measure of uncertainty is correct and around the original estimate.

```{r}
ci_spec_richness_natura2000_2 <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = bootstrap_results_natura2000,
  grouping_var = "year",
  type = "perc",
  conf = 0.95,
  no_bias = TRUE
)
ci_spec_richness_nonnatura2000_2 <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = bootstrap_results_nonnatura2000,
  grouping_var = "year",
  type = "perc",
  conf = 0.95,
  no_bias = TRUE
)
```

```{r}
#| warning: false
# Get bias values
bias_df_2 <- bind_rows(bootstrap_results_natura2000_df,
                     bootstrap_results_nonnatura2000_df) %>%
  distinct(year, estimate = est_original, `bootstrap estimate` = est_boot, location)

# Get estimate values
estimate_df_2 <- bias_df_2 %>%
  pivot_longer(cols = c("estimate", "bootstrap estimate"),
               names_to = "Legend", values_to = "value") %>%
  mutate(Legend = factor(Legend, levels = c("estimate", "bootstrap estimate"),
                         ordered = TRUE))
# Visualise
bind_rows(bootstrap_results_natura2000_df,
          bootstrap_results_nonnatura2000_df) %>%
  ggplot(aes(x = year)) +
  # Distribution
  geom_violin(aes(y = rep_boot, group = year),
              fill = alpha("cornflowerblue", 0.2)) +
  # Estimates and bias
  geom_point(data = estimate_df,
             aes(y = value, shape = Legend, colour = location),
             size = 1, alpha = 0.5) +
  # Intervals
  geom_errorbar(
    data = bind_rows(
      ci_spec_richness_natura2000_2 %>%
        mutate(location = "Natura2000",
               year = as.numeric(year)),
      ci_spec_richness_nonnatura2000_2 %>%
        mutate(location = "non-Natura2000",
               year = as.numeric(year))
    ),
    aes(ymin = ll, ymax = ul, colour = location),
    linewidth = 0.5
  ) +
  # Settings
  facet_wrap(~location, ncol = 1) +
  labs(y = "Species richness",
       x = "", shape = "Legend:", colour = "Location:") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

There are some issues with this method:

- The bootstrap should resample grid cells, not occurrence records
- The cubes have a different number of rows
- The bootstrap bias is very large

# Species richness density

This calculates average local richness.

> Are Natura 2000 cells richer in species than non-Natura 2000 cells?

- No need to adjust for area (every observation is one 1 km² cell)
- Easy to interpret (difference of means)
- Naturally accommodates different numbers of cells in the two habitats

## Bootstrapping

Because we have two datasets of different sizes, we cannot use **dubicube** for bootstrapping. (Perhaps via stratified bootstrapping with **boot** package?)

```{r}
# Combine datasets
cell_richness <- bind_rows(
  cube_natura2000$data %>%
    mutate(location = "Natura2000"),
  cube_nonnatura2000$data %>%
    mutate(location = "non-Natura2000")
) %>%
  summarise(
    richness = n_distinct(taxonKey),
    total_occ = sum(obs),
    .by = c("year", "cellCode", "location")
  ) %>%
  mutate(
    tot_occ_year_location = sum(total_occ),
    .by = c("year", "location")
  ) %>%
  mutate(
    tot_occ_year = sum(total_occ),
    .by = "year"
  ) %>%
  select(-"total_occ")
```

```{r}
# Number of bootstrap replicates
n_boot <- 1000

years <- sort(unique(cell_richness$year))

# Calculate species richness density per year
out <- purrr::map_dfr(years, function(yr) {
  # Filter year
  dat_year <- filter(cell_richness, year == yr)
  
  # Split data
  nat <- filter(dat_year, location == "Natura2000")
  nonnat <- filter(dat_year, location == "non-Natura2000")

  # Original estimate
  est_original <-
    mean(nat$richness) -
    mean(nonnat$richness)

  # Bootstrap replicates
  rep_boot <- replicate(n_boot, {
    nat_boot <- nat[sample(nrow(nat), replace = TRUE), ]
    nonnat_boot <- nonnat[sample(nrow(nonnat), replace = TRUE), ]

    mean(nat_boot$richness) -
      mean(nonnat_boot$richness)
  })
  
  # Calculate statistics
  est_boot <- mean(rep_boot)
  se_boot <- sd(rep_boot)
  bias_boot <- est_boot - est_original
  
  # Return results in dubicube format
  tibble(
    sample = seq_len(n_boot),
    year = yr,
    est_original = est_original,
    rep_boot = rep_boot,
    est_boot = est_boot,
    se_boot = se_boot,
    bias_boot = bias_boot
  )
})
```

We can calculate the intervals with **dubicube** (not BCa).

```{r}
# Calculate percentile intervals
ci_difference <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = out,
  grouping_var = "year",
  type = "perc",
  conf = 0.95,
  aggregate = FALSE
)
```

```{r}
threshold_specs <- 5
```

We classify effects with **dubicube**.
We use an arbitrary difference of `r threshold_specs` species to differentiate between large or small effects.

```{r}
# Perform effect classification
result_ci_difference <- dubicube::add_effect_classification(
  df = ci_difference,
  cl_columns = c("ll", "ul"),
  threshold = threshold_specs,
  reference = 0,
  coarse = TRUE
)
```

```{r}
# Define ggplot colour palettes
coarse_colour <- scale_fill_manual(
  values = c(
    "increase" = "chartreuse3",
    "stable"   = "gold",
    "decrease" = "firebrick1",
    "unknown"  = "skyblue"
  ),
  labels = c(
    increase = "Higher in Natura 2000",
    stable   = "No meaningful difference",
    decrease = "Lower in Natura 2000",
    unknown  = "Uncertain"
  ),
  drop = FALSE,
  name = NULL
)
```

```{r}
smooth_df <- result_ci_difference %>%
  distinct(year, est_original, ll, ul, effect_coarse)

p_boot_richness <- result_ci_difference %>%
  ggplot(aes(x = factor(year))) +

  # zero reference line
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # optional threshold lines
  geom_hline(
    yintercept = c(-threshold_specs, threshold_specs),
    linetype = "dotdash",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # CI bounds (raw)
  geom_errorbar(
    data = smooth_df,
    aes(ymin = ll, ymax = ul),
    linewidth = 0.6,
    width = 0.2,
    colour = "black",
    show.legend = TRUE
  ) +

  # point estimate
  geom_point(
    data = smooth_df,
    aes(y = est_original, fill = effect_coarse),
    size = 2,
    shape = 21,
    colour = "black",
    stroke = 0.3,
    show.legend = TRUE
  ) +

  # smoother: estimate trend
  geom_smooth(
    data = smooth_df,
    aes(y = est_original, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "darkorange",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI lower bound
  geom_smooth(
    data = smooth_df,
    aes(y = ll, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI upper bound
  geom_smooth(
    data = smooth_df,
    aes(y = ul, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  coarse_colour +
  scale_x_discrete(
    breaks = seq(1975, 2024, by = 5) |> as.character()
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10)
  ) +

  labs(
    y = "Difference in species richness density",
    x = "Year"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.box = "vertical"
  )

save_figure(p_boot_richness, "sp_richness_boot_coarse", path = out_path,
            dpi = 300, width = 8, height = 6)
p_boot_richness
```

## Model based

We can also perform this analysis in a model based way (GLMM).
Remember we aggregated:

```{r}
head(cell_richness)
```

```{r}
cell_richness %>%
  ggplot(aes(x = richness)) +
  geom_histogram() +
  facet_wrap(~location)
```

A Poisson model will not be reasonable (variance is much higher than mean):

```{r}
cell_richness %>%
  summarise(mean = mean(richness),
            var = var(richness),
            .by = "location")
```

Although not ideal, a Gaussian model directly models the mean richness and produces contrasts that are easy to interpret. Then the model is

$$
Y_{ijt} \sim \mathcal{N}(\mu_{ijt}, \sigma^2)
$$

where

* $Y_{ijt}$ = observed local species richness in cell $i$, habitat $j$, year $t$
* $b_i$ = random intercept for cell $i$
* $\mu_{ijt} = \mathbb{E}[Y_{ijt}]$ = expected richness

with linear predictor

$$
\mu_{ijt} = \beta_0 + \beta_{\mathrm{loc}(j)} + \beta_{\mathrm{year}(t)} + \beta_{\mathrm{loc}\times\mathrm{year}(j,t)} + b_i
$$

where

$$
b_i \sim \mathcal{N}(0,\sigma_b^2).
$$

The interaction term $\beta_{\mathrm{loc}\times\mathrm{year}(j,t)}$ allows the difference between Natura 2000 and non-Natura 2000 to vary from year to year.

```{r}
# Create factors
cell_richness_f <- cell_richness %>%
  mutate(
    year_f = factor(year),
    cellCode_f = factor(cellCode),
    location_f = factor(location, levels = c("non-Natura2000", "Natura2000"))
  )
```

```{r}
fit_richness <- cache_compute(
  path = cache_path,
  file = "fit_richness",
  expr = {
    glmmTMB(
      richness ~ location_f*year_f + (1 | cellCode_f),
      data = cell_richness_f
    )
  }
)
```

```{r}
# Calculate contrasts and confidence limits
emm <- emmeans(fit_richness, ~ location_f | year_f)
diff_year <- contrast(
  emm,
  method = "revpairwise",
  by = "year_f"
)
diff_df <- as.data.frame(diff_year)

diff_df <- diff_df %>%
  mutate(
    ll = estimate - 1.96 * SE,
    ul = estimate + 1.96 * SE
  )
```

```{r}
# Perform effect classification
smooth_df_model <- dubicube::add_effect_classification(
  df = diff_df,
  cl_columns = c("ll", "ul"),
  threshold = threshold_specs,
  reference = 0,
  coarse = TRUE
) %>%
  mutate(year = as.numeric(year_f))

p_model_richness <- smooth_df_model %>%
  ggplot(aes(x = year_f)) +

  # zero line
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # threshold lines
  geom_hline(
    yintercept = c(-threshold_specs, threshold_specs),
    linetype = "dotdash",
    linewidth = 0.5,
    colour = "grey40"
  ) +
  
  # CI bounds (raw)
  geom_errorbar(
    aes(ymin = ll, ymax = ul),
    linewidth = 0.6,
    width = 0.2,
    colour = "black",
    show.legend = TRUE
  ) +

  # point estimate
  geom_point(
    aes(y = estimate, fill = effect_coarse),
    size = 2,
    shape = 21,
    colour = "black",
    stroke = 0.3,
    show.legend = TRUE
  ) +

  # smoother: estimate trend
  geom_smooth(
    aes(y = estimate, x = year),
    method = "loess",
    se = FALSE,
    colour = "darkorange",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI lower bound
  geom_smooth(
    aes(y = ll, x = year),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI upper bound
  geom_smooth(
    aes(y = ul, x = year),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  coarse_colour +
  scale_x_discrete(
    breaks = seq(1975, 2024, by = 5) |> as.character()
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10)
  ) +

  labs(
    y = "Difference in species richness density",
    x = "Year"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.box = "vertical"
  )

save_figure(p_model_richness, "sp_richness_model_coarse", path = out_path,
            dpi = 300, width = 8, height = 6)
p_model_richness
```

```{r}
# Determine common y-axis limits from both input datasets
y_limits <- range(
  c(
    smooth_df$ll,
    smooth_df$ul,
    smooth_df_model$ll,
    smooth_df_model$ul
  ),
  na.rm = TRUE
)

# Apply common limits and add titles
p_boot_richness_final <-
  p_boot_richness +
  coord_cartesian(ylim = y_limits) +
  ggtitle("Bootstrap") +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10)
  )

p_model_richness_final <-
  p_model_richness +
  coord_cartesian(ylim = y_limits) +
  ggtitle("Model-based") +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10)
  )

# Combine
p_combined <- plot_grid(
  p_boot_richness_final,
  p_model_richness_final,
  labels = c("A.", "B."),
  ncol = 1,
  align = "hv",
  axis = "tblr"
)

save_figure(p_combined, "sp_richness_coarse", path = out_path,
            dpi = 300, width = 8, height = 8)
```


## Sampling effort (model based)
We attempted to include survey effort in the model using the log-transformed total number of occurrences per habitat and year.

However, this approach does not work well in practice. The effort variable is strongly confounded with the year × habitat structure in the data, meaning it is not independent from the variables already in the model. As a result, the model becomes rank-deficient and cannot uniquely estimate the effect of effort: some coefficients are dropped because they are not identifiable.

In short, we tried to account for sampling effort, but the available effort proxy is too closely tied to temporal and habitat variation to be estimated separately from the main effects.

# Occurrence density
## Bootstrapping

We use the same method as before.

```{r}
# Combine datasets
cell_occurrences <- bind_rows(
  cube_natura2000$data %>%
    mutate(location = "Natura2000"),
  cube_nonnatura2000$data %>%
    mutate(location = "non-Natura2000")
) %>%
  summarise(
    total_occ = sum(obs),
    .by = c("year", "cellCode", "location")
  )
```

```{r}
# Number of bootstrap replicates
n_boot <- 1000

years <- sort(unique(cell_occurrences$year))

# Calculate species richness density per year
out_occurrences <- purrr::map_dfr(years, function(yr) {
  # Filter year
  dat_year <- filter(cell_occurrences, year == yr)
  
  # Split data
  nat <- filter(dat_year, location == "Natura2000")
  nonnat <- filter(dat_year, location == "non-Natura2000")

  # Original estimate
  est_original <-
    mean(nat$total_occ) -
    mean(nonnat$total_occ)

  # Bootstrap replicates
  rep_boot <- replicate(n_boot, {
    nat_boot <- nat[sample(nrow(nat), replace = TRUE), ]
    nonnat_boot <- nonnat[sample(nrow(nonnat), replace = TRUE), ]

    mean(nat_boot$total_occ) -
      mean(nonnat_boot$total_occ)
  })
  
  # Calculate statistics
  est_boot <- mean(rep_boot)
  se_boot <- sd(rep_boot)
  bias_boot <- est_boot - est_original
  
  # Return results in dubicube format
  tibble(
    sample = seq_len(n_boot),
    year = yr,
    est_original = est_original,
    rep_boot = rep_boot,
    est_boot = est_boot,
    se_boot = se_boot,
    bias_boot = bias_boot
  )
})
```

We can calculate the intervals with **dubicube** (not BCa).

```{r}
# Calculate percentile intervals
ci_difference_occurrences <- dubicube::calculate_bootstrap_ci(
  bootstrap_results = out_occurrences,
  grouping_var = "year",
  type = "perc",
  conf = 0.95,
  aggregate = FALSE
)
```

```{r}
threshold_occs <- 15
```

We classify effects with **dubicube**.
We use an arbitrary difference of `r threshold_specs` species to differentiate between large or small effects.

```{r}
# Perform effect classification
result_ci_difference_occurrences <- dubicube::add_effect_classification(
  df = ci_difference_occurrences,
  cl_columns = c("ll", "ul"),
  threshold = threshold_occs,
  reference = 0,
  coarse = TRUE
)
```

```{r}
smooth_df_occurrences <- result_ci_difference_occurrences %>%
  distinct(year, est_original, ll, ul, effect_coarse)

p_boot_occurrences <- result_ci_difference_occurrences %>%
  ggplot(aes(x = factor(year))) +

  # zero reference line
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # optional threshold lines
  geom_hline(
    yintercept = c(-threshold_occs, threshold_occs),
    linetype = "dotdash",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # CI bounds (raw)
  geom_errorbar(
    data = smooth_df_occurrences,
    aes(ymin = ll, ymax = ul),
    linewidth = 0.6,
    width = 0.2,
    colour = "black",
    show.legend = TRUE
  ) +

  # point estimate
  geom_point(
    data = smooth_df_occurrences,
    aes(y = est_original, fill = effect_coarse),
    size = 2,
    shape = 21,
    colour = "black",
    stroke = 0.3,
    show.legend = TRUE
  ) +

  # smoother: estimate trend
  geom_smooth(
    data = smooth_df_occurrences,
    aes(y = est_original, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "darkorange",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI lower bound
  geom_smooth(
    data = smooth_df_occurrences,
    aes(y = ll, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI upper bound
  geom_smooth(
    data = smooth_df_occurrences,
    aes(y = ul, x = as.numeric(factor(year))),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  coarse_colour +
  scale_x_discrete(
    breaks = seq(1975, 2024, by = 5) |> as.character()
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10)
  ) +

  labs(
    y = "Difference in occurrence density",
    x = "Year"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.box = "vertical"
  )

save_figure(p_boot_occurrences, "sp_occs_boot_coarse", path = out_path,
            dpi = 300, width = 8, height = 6)
p_boot_occurrences
```

## Model based

We can perform this analysis in a model based way like before.

```{r}
# Create factors
cell_occurrences_f <- cell_occurrences %>%
  mutate(
    year_f = factor(year),
    cellCode_f = factor(cellCode),
    location_f = factor(location, levels = c("non-Natura2000", "Natura2000"))
  )
```

```{r}
fit_occurrences <- cache_compute(
  path = cache_path,
  file = "fit_occurrences",
  expr = {
    glmmTMB(
      total_occ ~ location_f*year_f + (1 | cellCode_f),
      data = cell_occurrences_f
    )
  }
)
```

```{r}
# Calculate contrasts and confidence limits
emm_occurrences <- emmeans(fit_occurrences, ~ location_f | year_f)
diff_year_occurrences <- contrast(
  emm_occurrences,
  method = "revpairwise",
  by = "year_f"
)
diff_df_occurrences <- as.data.frame(diff_year_occurrences)

diff_df_occurrences <- diff_df_occurrences %>%
  mutate(
    ll = estimate - 1.96 * SE,
    ul = estimate + 1.96 * SE
  )
```

```{r}
# Perform effect classification
smooth_df_model_occurrences <- dubicube::add_effect_classification(
  df = diff_df_occurrences,
  cl_columns = c("ll", "ul"),
  threshold = threshold_occs,
  reference = 0,
  coarse = TRUE
) %>%
  mutate(year = as.numeric(year_f))

p_model_occurrences <- smooth_df_model_occurrences %>%
  ggplot(aes(x = year_f)) +

  # zero line
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey40"
  ) +

  # threshold lines
  geom_hline(
    yintercept = c(-threshold_specs, threshold_specs),
    linetype = "dotdash",
    linewidth = 0.5,
    colour = "grey40"
  ) +
  
  # CI bounds (raw)
  geom_errorbar(
    aes(ymin = ll, ymax = ul),
    linewidth = 0.6,
    width = 0.2,
    colour = "black",
    show.legend = TRUE
  ) +

  # point estimate
  geom_point(
    aes(y = estimate, fill = effect_coarse),
    size = 2,
    shape = 21,
    colour = "black",
    stroke = 0.3,
    show.legend = TRUE
  ) +

  # smoother: estimate trend
  geom_smooth(
    aes(y = estimate, x = year),
    method = "loess",
    se = FALSE,
    colour = "darkorange",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI lower bound
  geom_smooth(
    aes(y = ll, x = year),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  # smoother: CI upper bound
  geom_smooth(
    aes(y = ul, x = year),
    method = "loess",
    se = FALSE,
    colour = "orange",
    linetype = "dotted",
    linewidth = 1,
    formula = "y ~ x"
  ) +

  coarse_colour +
  scale_x_discrete(
    breaks = seq(1975, 2024, by = 5) |> as.character()
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10)
  ) +

  labs(
    y = "Difference in occurrence density",
    x = "Year"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.box = "vertical"
  )

save_figure(p_model_occurrences, "sp_occs_model_coarse", path = out_path,
            dpi = 300, width = 8, height = 6)
p_model_occurrences
```

```{r}
# Determine common y-axis limits from both input datasets
y_limits_occurrences <- range(
  c(
    smooth_df_occurrences$ll,
    smooth_df_occurrences$ul,
    smooth_df_model_occurrences$ll,
    smooth_df_model_occurrences$ul
  ),
  na.rm = TRUE
)

# Apply common limits and add titles
p_boot_occurrences_final <-
  p_boot_occurrences +
  coord_cartesian(ylim = y_limits_occurrences) +
  ggtitle("Bootstrap") +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10)
  )

p_model_occurrences_final <-
  p_model_occurrences +
  coord_cartesian(ylim = y_limits_occurrences) +
  ggtitle("Model-based") +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10)
  )

# Combine
p_combined_occurrences <- plot_grid(
  p_boot_occurrences_final,
  p_model_occurrences_final,
  labels = c("A.", "B."),
  ncol = 1,
  align = "hv",
  axis = "tblr"
)

save_figure(p_combined_occurrences, "sp_occs_coarse", path = out_path,
            dpi = 300, width = 8, height = 8)
```
