---
title: "B3_filter_angio"
author: "Toon Van Daele"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
remotes::install_github("b-cubed-eu/dubicube")
library(dubicube)
library(dplyr)
library(ggplot2)
```

## load cube

```{r }
angioGBIF <- read.csv("./_preprocessing/data/raw/Angiosperm_data_Flanders_1km.csv")
```

### check for zero's and NoData

```{r}
angioGBIF %>%
  filter(occurrences == 0)

angioGBIF %>%
  filter(is.na(occurrences))

```

No NA's, no zero occurrences

## Pre-process Cube

```{r }
angio_p <- b3gbi::process_cube(angioGBIF)
```

## general check with dubicube

```{r}
diagnose_cube(angio_p)
```

```{r}
basic_cube_rules()
```

```{r}
rule_spatial_min_cells()
```

# Check distribution of observations among cells

## total occurences

```{r}
table(is.na(angio_p$data$obs))
```

```{r}
angio_p$data %>%
  group_by(xcoord, ycoord, cellCode) %>%
  summarize(sum_obs = sum(obs)) %>%
  ggplot(aes(x = xcoord, y = ycoord, colour = sum_obs)) + geom_point() +
  coord_equal()
```

The locations with outliers or nodata are not visible because the points inµ the plot overlap.

## Zoom to problematic region

```{r}
angio_p$data %>%
  group_by(xcoord, ycoord, cellCode) %>%
  summarize(sum_obs = sum(obs)) %>%
  dplyr::filter(xcoord > 3820000 & xcoord < 3860000 &
                ycoord > 3105000 & ycoord < 3121000) %>%
  ggplot(aes(x = xcoord, y = ycoord, colour = sum_obs)) + geom_point()
```

```{r}
angio_p$data %>%
  group_by(xcoord, ycoord, cellCode) %>%
  summarize(sum_obs = sum(obs)) %>%
  dplyr::filter(xcoord > 3820000 & xcoord < 3860000 &
                ycoord > 3105000 & ycoord < 3121000)
```

The eea gridcells with 0 occurences (or no data) are not exported? These locations are not outliers, nor NA. The grid locations are not exported. -\> Seems to be an issue for the GBIF export.

## Distribution of total observations by gridcell

```{r}
angio_p$data %>%
  group_by(cellCode) %>%
  summarize(`Number of observations` = sum(obs)) %>%
  ggplot(aes(x = `Number of observations`)) +
  geom_histogram(fill = "#005957", color = NA) +
  theme_minimal() +
  theme(
    axis.title = element_text(),
    axis.text = element_text()
  )
```

It's very strong skewed to the right, to be expected for a variable 'number of observations'.

```{r}
angio_p_log <- angio_p$data %>%
  group_by(cellCode, xcoord, ycoord) %>%
  summarize(log_sum_obs = log(sum(obs)), .groups = "drop")
```

Log transformation -\> in the log scale the variable is normal distributed

red = normal distribution black = density function of the actual distribution in the log scale)

```{r}
ggplot(angio_p_log, aes(x = log_sum_obs)) + 
  geom_density() +
  stat_function(fun = dnorm, colour = "#48A529",
                args = list(mean = mean(angio_p_log$log_sum_obs),
                            sd = sd(angio_p_log$log_sum_obs)))+
  labs(
    x = "Log(Number of occurrences)",
    y = "Frequency"
  )+
  theme_minimal()
```

Solution for visualization -\> a legend in the log scale?

# option 1: linear legend, but adapted color gradient

```{r}
df_plot <- angio_p$data %>%
  group_by(xcoord, ycoord, cellCode) %>%
  summarize(sum_obs = sum(obs))

ggplot(df_plot, aes(x = xcoord, y = ycoord, colour = sum_obs)) + geom_point() +
  coord_equal() +
  scale_colour_viridis_c(option = "D",
                         values = c(0, 0.01, 0.05, 0.1, 1))
# values can be adjusted manually
```

# option 2: legend in the log scale

```{r}
ggplot(df_plot, aes(x = xcoord, y = ycoord, colour = sum_obs)) + geom_point() +
  coord_equal() +
  scale_colour_viridis_c(option = "D",
                         transform = "log10",
                         breaks = c(1, 10, 100, 1000, 10000))
```

## Are there outliers in the log scale?

-\> IQR \> 1.5

```{r}
ggplot(angio_p_log, aes(y = log_sum_obs)) +
  geom_boxplot()
```

Location on the map:

```{r}
#IQR(angio_p_log$log_sum_obs)

angio_stats <- boxplot.stats(angio_p_log$log_sum_obs)

angio_outl <- angio_stats$out


angio_p_log <- angio_p_log %>%
  mutate(outl = case_when(log_sum_obs < angio_stats$stats[1] ~ "lower",
                          log_sum_obs > angio_stats$stats[5] ~ "upper",
                          .default = NA))

ggplot(angio_p_log, aes(x = xcoord, y = ycoord)) + geom_point(aes(colour = log_sum_obs)) +
  coord_equal() +
  scale_colour_viridis_c(option = "D") +
  geom_point(data = angio_p_log %>%
               dplyr::filter(outl %in% c("upper", "lower")),
             aes(shape = outl))

# Maybe the MAD (mean absolute deviation) is a better alternative (to be cheked)
```

Still outliers, but not so many

```{r}
table(angio_p_log$outl)
```

## Possible workflows for a more generic check for outliers

use a gamma distribution -\> more flexible (to be checked)

- fit a gamma model (group by op cellen of op cellen+year)
- get shape & scale values
- transform data with the alpha parameter (Wilson Hilferty transformation)
- IQR boxplot \> 1.5 of 3 of onder bepaalde probability threshold p \< 0.0001
- ofwel vernieuwende aanpak: <https://www.mdpi.com/2075-1680/12/2/107>

# Check publication time-lag

```{r}
ts <- angio_p$data %>%
  group_by(year) %>%
  summarize(sum_obs = sum(obs))

gam_ts <- mgcv::gam(sum_obs ~ s(year), data = ts)

plot(gam_ts)
```
