---

editor: 
  markdown: 
    wrap: 72
---

```{r}
#install packages

#install.packages("remotes")
#remotes::install_github("b-cubed-eu/invasimapr")

#suppressPackageStartupMessages({
#    need = c("dplyr","tidyr","purrr","tibble","stringr","ggplot2","Matrix","glmnet#","rlang", "glmmTMB",
#    "performance", "vegan", "cluster", "matrixStats", "sf", "MASS", "factoextra", #"viridis", "pheatmap")
#    to_install = need[!sapply(need, requireNamespace, quietly = TRUE)]
#     if (length(to_install)) install.packages(to_install)
#     lapply(need, library, character.only = TRUE)
#   })

deps = c("dplyr","tidyr","purrr","tibble","stringr", "forcats",
  "ggplot2",
  # Modelling / stats
  "glmmTMB","performance","Matrix","glmnet","MASS","vegan","cluster","matrixStats",
  # Spatial
  "sf",
  # Viz helpers used in later steps
  "viridis","pheatmap","factoextra")
  
library(rlang)

  is_installed = vapply(deps, requireNamespace, logical(1), quietly = TRUE)
missing = deps[!is_installed]

cat("### Dependency check\n\n")
if (length(missing) == 0) {
  cat("- All recommended packages are available.\n")
} else {
  cat("- Missing packages (install before running the full workflow):\n")
  cat(paste0("  - ", missing, collapse = "\n"), "\n")
}
)

```

## Data access and preparation

```{r}
# Ensure the package is loaded
 install.packages("remotes")
 remotes::install_github("b-cubed-eu/dissmapr")

# Ensure the package is loaded
if (!requireNamespace("dissmapr", quietly = TRUE)) {
  knitr::knit_exit()
}
# Optional: report package version
packageVersion("dissmapr")
```

```{r}
angio_data = dissmapr::get_occurrence_data(
    source_type = "gbif",
  gbif_zip_url= "https://api.gbif.org/v1/occurrence/download/request/0049575-260519110011954.zip"

)



```

```{r}
library(dplyr)

angio_data <- angio_data%>%
  filter(!is.na(y))%>%
  filter(!is.na(year))%>%
  filter(!is.na(verbatimScientificName))%>%
  filter(!is.na(pa))

angio_data_test <- angio_data[1:1000000,]

angio_result = dissmapr::format_df(
  data        = angio_data_test, # A `data.frame` of biodiversity records
  species_col = "verbatimScientificName", # Name of species column (required for `"long"`)
  value_col   = "pa", # Name of value column (e.g. presence/abundance; for `"long"`)
  extra_cols  = NULL, # Character vector of other columns to keep
  format      = "long" # Either`"long"` or `"wide"`
)


# Check `bfly_result` structure
str(angio_result, max.level = 1)

# Optional: Create new objects from list items
site_obs = angio_result$site_obs
site_spp = angio_result$site_spp

# Check results
dim(site_obs)
head(site_obs)

dim(site_spp)
head(site_spp[, 1:6])

#### Get parameters from processed data to use later
# Number of species
(n_sp = dim(site_spp)[2] - 3)

# Species names
sp_cols = names(site_spp)[-c(1:3)]
sp_cols[1:10]
```

```{r}
# Load the national boundary
rsa = sf::st_read('./_preprocessing/data/raw/spatial/flanders_noholes.geojson')

# Choose a working resolution
res = 0.1 # decimal degrees° (≈ 55 km at the equator)

# Convert the AoI to a 'terra' vector
rsa_vect = terra::vect(rsa)

# Initialise a blank raster template
grid = terra::rast(rsa_vect, resolution = res, crs = terra::crs(rsa_vect))

# Populate the raster with placeholder values
terra::values(grid) = 1

# Clip the raster to the AoI
grid_masked = terra::mask(grid, rsa_vect)

# Generate a 0.5° grid summary for the point dataset `site_spp`
grid_list = dissmapr::generate_grid(
  data          = site_spp, # point data with x/y + species columns
  x_col         = "x", # longitude column
  y_col         = "y", # latitude  column
  grid_size     = 0.5, # cell size in degrees
  sum_cols      = names(site_spp)[4:ncol(site_spp)],
  # columns to aggregate * could also use `names(site_spp)[4:ncol(site_spp)]`
  crs_epsg      = 4326 # WGS84
)

# Inspect the returned list
str(grid_list, max.level = 1)

# (Optional) Promote list items to named objects
grid_r = grid_list$grid_r$grid_id # raster
grid_sf = grid_list$grid_sf # polygons for mapping or joins
grid_spp = grid_list$grid_spp # tabular summary per cell
grid_spp_pa = grid_list$grid_spp_pa # presence/absence summary

# Quick checks
dim(grid_sf) # ; head(grid_sf)
dim(grid_spp) # ; head(grid_spp[, 1:8])
dim(grid_spp_pa) # ; head(grid_spp_pa[, 1:8])

```
```{r}
# Read in target species list
species = read.csv(system.file("extdata",
  "rsa_butterfly_species_names_n27_100plus.csv",
  package = "invasimapr"
), stringsAsFactors = FALSE)$species

# Filter `grid_spp` and convert to long-format
grid_obs = grid_spp |>
  dplyr::select(-mapsheet) |> # Drop mapsheet metadata
  tidyr::pivot_longer(
    cols = -c(grid_id, centroid_lon, centroid_lat, obs_sum, spp_rich), # Keep core metadata columns only
    names_to = "species",
    values_to = "count",
    values_drop_na = TRUE
  ) |>
  dplyr::filter(
    # obs_sum > 100, # Only high-observation sites
    count > 0,# Remove absent species
    .data$species %in% .env$species# Keep only target species
    ) |>
  dplyr::rename(
    site_id = grid_id, # Change 'grid_id' to 'site_id'
    x = centroid_lon, # Change 'centroid_lon' to 'x'
    y = centroid_lat # Change 'centroid_lat' to 'y'
  ) |>
  dplyr::relocate(site_id, x, y, obs_sum, spp_rich, species, count)

dim(grid_obs)
head(grid_obs)
length(unique(grid_obs$species))
length(unique(grid_obs$site))

# Reshape site-by-species matrix to wide format and clean
grid_spp = grid_obs |>
  tidyr::pivot_wider(
    names_from = species,
    values_from = count,
    values_fill = 0 # Replace missing counts with 0
  )

dim(grid_spp)
# head(grid_spp)
```


```{r}
# Retrieve 19 bioclim layers (≈10-km, WorldClim v2.1) for all grid centroids
data_path = "_data" # cache folder for rasters
enviro_list = dissmapr::get_enviro_data(
  data       = grid_spp, # centroids + obs_sum + spp_rich
  buffer_km  = 10, # pad the AOI slightly
  source     = "geodata", # WorldClim/SoilGrids interface
  var        = "bio", # bioclim variable set
  res        = 5, # 5-arc-min ≈ 10 km
  grid_r     = grid_r, # To set resampling resolution, if necessary
  path       = data_path,
  sp_cols    = 7:ncol(grid_spp), # ignore species columns
  ext_cols   = c("obs_sum", "spp_rich") # carry effort & richness through
)

# Quick checks
str(enviro_list, max.level = 1)

# (Optional) Assign concise layer names for readability
# Find names here https://www.worldclim.org/data/bioclim.html
names_env = c(
  "temp_mean", "mdr", "iso", "temp_sea", "temp_max", "temp_min",
  "temp_range", "temp_wetQ", "temp_dryQ", "temp_warmQ",
  "temp_coldQ", "rain_mean", "rain_wet", "rain_dry",
  "rain_sea", "rain_wetQ", "rain_dryQ", "rain_warmQ", "rain_coldQ"
)
names(enviro_list$env_rast) = names_env

# (Optional) Promote frequently-used objects
env_r = enviro_list$env_rast # cropped climate stack
env_df = enviro_list$env_df # site × environment data-frame

# Quick checks
env_r
dim(env_df)
head(env_df)

# Build the final site × environment table
grid_env = env_df |>
  dplyr::select(
    site_id, x, y,
    obs_sum, spp_rich, dplyr::everything()
  ) |>
  dplyr::mutate(across(
    .cols = -c(site_id, x, y, obs_sum, spp_rich), # all other columns
    .fns = ~ as.numeric(scale(.x)), # Scale bio
    .names = "{.col}" # keep same names
  ))

str(grid_env, max.level = 1)
head(grid_env)
```

```{r}
# # (Optional) Rename BIO
# names(env_df) = c("grid_id", "centroid_lon", "centroid_lat", names_env, "obs_sum", "spp_rich")
#
# # Run the filter and compare dimensions
# # Filter environmental predictors for |r| > 0.70
# env_vars_reduced = dissmapr::rm_correlated(
#   data       = env_df[, 4:23],  # drop ID + coord columns
#   cols       = NULL,                  # infer all numeric cols
#   threshold  = 0.70,
#   plot       = TRUE                   # show heat-map of retained vars
# )
#
# # Before vs after
# c(original = ncol(env_df[, c(4, 6:24)]),
#   reduced  = ncol(env_vars_reduced))
```

### Data access and preparation using `invasimapr`

This section sets up data needed by all downstream wrappers: site locations, environmental covariates, resident community matrices, and species traits. It also shows how to enrich traits with metadata, and how to simulate invader trait tables for later steps.

#### Retrieve and link trait and metadata for each species

`get_trait_data()` provides an automated pipeline for extracting and joining both biological trait data and rich metadata for any focal species. Many analyses benefit from curated trait columns, short taxonomy summaries, and links to images. `get_trait_data()` consolidates these per species from local trait tables and optional online sources, returning a single tidy row per species for frictionless joins.

:hourglass_flowing_sand: **What it does**: Given a species name, the function looks up traits in a local table (or [TRY](https://www.try-db.org/TryWeb/Home.php)-style database), performs tolerant name matching, optionally scrapes a short [Wikipedia](https://www.wikipedia.org/) description and taxonomy, and can extract a compact image-based colour palette for visual summaries, as follows:

1.  **Trait Table Lookup**: Retrieves species' trait data from a local trait table (CSV) or a [TRY](https://www.try-db.org/TryWeb/Home.php) (Kattge et al 2020) style database, using fuzzy matching to ensure robust linkage even when there are minor naming inconsistencies.
2.  **Wikipedia Metadata Scraping**: Optionally augments each species entry with a taxonomic summary, higher taxonomy, and representative images scraped directly from [Wikipedia](https://www.wikipedia.org/).
3.  **Image-based Color Palette Extraction**: If enabled, downloads and processes public domain images to extract the most frequent colors, optionally removing green/white backgrounds to focus on diagnostic features.
4.  **Flexible Output**: Returns a single-row tibble with the species name, trait data, taxonomic metadata, image URL, and color palette - all harmonized for downstream analyses or visualization.

:information_source: **Why this matters**: Having a consistent, analysis-ready trait table avoids downstream NA cascades, eases plotting, and ensures that residents and invaders share identical column definitions.

:warning: **Practical checks and tips**: Keep resident and invader trait column names identical; prefer factors for categorical traits with explicit levels; store provenance for any scraped fields for reproducibility.

\`\`\`{r get-traits} \# Fetch local trait data.frame btfly_traits = read.csv(system.file("extdata", "species_traits.csv", package = "invasimapr"))

str(btfly_traits)

# Retrieve and join trait/metadata for all species in the observation set

spp_traits = purrr::map_dfr( unique(grid_obs$species), # unique(longDF$species), \~ invasimapr::get_trait_data( species = .x, n_palette = 5, preview = FALSE, do_summary = TRUE, do_taxonomy = TRUE, do_image = TRUE, do_palette = TRUE, local_trait_df = btfly_traits, local_species_col = "species", max_dist = 1 ) ) \# The final output combines trait data, taxonomic info, Wikipedia summary, images, and color palette for each species. \# This integrated dataset supports multi-faceted biodiversity, trait, and visualization analyses.

# Check output

str(spp_traits, 1) head(spp_traits\[1:5,1:5\])
