---
title: "Calculate General Indicators Flanders Case Study WP6"
output: html_document
---

#install packages

```{r}
#install.packages("b3gbi", repos = c("https://b-cubed-eu.r-universe.dev", "https://cloud.r-project.org"))
#install.packages("rnaturalearthdata")
#install.packages("sf")
#install.packages("taxize")
#install.packages("here")


gc()
library(b3gbi)
library(sf)
library(taxize)
library(here)
```

#Define input file and directory of output files

```{r}
#more info about download here (focussing on dataset of Florabank): https://www.gbif.org/occurrence/download/0051756-260226173443078
download_key <- "0051756-260226173443078"
#download data
dl <- occ_download_get(download_key, overwrite = TRUE)
#import data
dat <- occ_download_import(dl)


#only data before 2023 is complete in GBIF for Florabank
data <- dat%>%filter(year<2023)

#only maintain cells which are well surveyed according to methodology of Florabank (per year and per 1km gridcell more than 100 unique species observed)
data_filtered <- data %>%
  group_by(eeacellcode, year) %>%            # groep per cel en jaar
  filter(n_distinct(species) > 100) %>%      # enkel groepen met >100 unieke soorten
  ungroup() 



Focus <- 'Flanders_Florabank_1km_well_surveyed_cells'


output_path <- here("_preprocessing","data", "output", "general_indicators", Focus)

if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
}
#check resolution of file
Angiosperm_data_cube <- process_cube(data_filtered)

str(Angiosperm_data_cube)
```

#Running b3gbi indicators ##8.1. Observed Species Richness \~ recorded richness

```{r}
'Observed richness is calculated by summing the number of unique species observed for each year or each cell. Observed richness is highly dependent on the comprehensiveness of the dataset it is being applied to. If some regions are more intensively, carefully or systematically sampled than others, this will likely reflect as higher observed richness. Observed richness also depends on the relative abundance and spatial aggregation of each species, with less abundant and less aggregated species less likely to be discovered during surveys (Hillebrand et al., 2018), as well as the detectability of each species.'

OSR_map <- obs_richness_map(Angiosperm_data_cube,
                            cell_size="grid")

saveRDS(OSR_map, 
        paste0(output_path, '/OSR_map.rds'))

OSR_ts <- obs_richness_ts(Angiosperm_data_cube, 
                          cell_size="grid")

saveRDS(OSR_ts, 
        paste0(output_path, '/OSR_ts.rds'))

```

##8.2. Total Occurrences

```{r}
'The total number of occurrences is calculated by summing the occurrences of all species observed for each cell or year. While not itself an indicator, this variable provides an overview of the comprehensiveness and distribution of data in the cube being analysed, and may be helpful, or even vital, for interpreting the results of calculated indicators.'

gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
TO_map <- total_occ_map(Angiosperm_data_cube,
                             cell_size="grid")



#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
TO_ts <- total_occ_ts(Angiosperm_data_cube, 
                           cell_size="grid")

saveRDS(TO_map, 
        paste0(output_path, '/TO_map.rds'))

saveRDS(TO_ts, 
        paste0(output_path, '/TO_ts.rds'))



#Bootstrapping ok
```


#8.4. Rarity

```{r}
"Rarity quantifies the scarcity or infrequency of species, and when summed over multiple species, serves as a crucial biodiversity indicator for conservation.A rare species might have a small population size, a limited distribution, or a unique ecological niche (Maciel, 2021; Rabinowitz, 1981). Rarity can also be a biodiversity indicator when summed over multiple species in an area, and may provide important insight for determining conservation priorities. When measured over time, rarity may indicate potential threats or changes in the environment. b3gbi offers two distinct measures: Abundance-Based Rarity (based on species' proportional occurrences) and Area-Based Rarity (based on species' spatial occupancy). These can be mapped (indicator_map) to identify areas with a higher presence of rare species, or tracked over time (indicator_ts) to observe changes in overall rarity."

#bootstrapping ok

gc()

ArRM_map <-  area_rarity_map(Angiosperm_data_cube,
                            cell_size="grid")

ArRM_map_plot <- plot(ArRM_map)

AbRM_map <- ab_rarity_map(Angiosperm_data_cube,
                            cell_size="grid")

AbRM_map_plot <- plot(AbRM_map)

ArRM_ts <-  area_rarity_ts(Angiosperm_data_cube,
                            cell_size="grid")

ArRM_ts_plot <- plot(ArRM_ts)

AbRM_ts <- ab_rarity_ts(Angiosperm_data_cube,
                            cell_size="grid")

AbRM_ts_plot <- plot(AbRM_ts)

##export data to showcase in dashboard
saveRDS(ArRM_map, 
        paste0(output_path, '/arRM_map.rds'))

saveRDS(ArRM_ts, 
        paste0(output_path, '/arRM_ts.rds'))


saveRDS(AbRM_map, 
        paste0(output_path, '/abRM_map.rds'))

saveRDS(AbRM_ts, 
        paste0(output_path, '/abRM_ts.rds'))

```

#8.5. Estimated Hill Diversity

```{r}
'Hill Diversity provides a unified framework for various diversity measures, allowing for different emphases on rare versus common species through a single parameter, q. b3gbi calculates three common forms: q=0 (approximates Species Richness, weighing all species equally), q=1 (emphasizes common species, like Hill-Shannon), and q=2 (emphasizes very common species, like Hill-Simpson). These indicators represent the "effective number of species" and can be mapped (indicator_map) or tracked over time (indicator_ts) to provide multi-faceted insights into biodiversity patterns.

SR: Species richness - hill0_map() and hill0_ts()

HSh: Hill-Shannon diversity - hill1_map() and hill1_ts()

HSi: Hill-Simpson diversity - hill2_map() and hill2_ts()'

gc()

SR_map <-  hill0_map(Angiosperm_data_cube,
                            cell_size="grid")


#HSh_map <- hill1_map(Angiosperm_data_cube,
#                            cell_size="grid")


HSi_map <- hill2_map(Angiosperm_data_cube,
                            cell_size="grid")


SR_ts <-  hill0_ts(Angiosperm_data_cube,
                            cell_size="grid")


#HSh_ts <- hill1_ts(Angiosperm_data_cube,
#                            cell_size="grid")


HSi_ts <- hill2_ts(Angiosperm_data_cube,
                            cell_size="grid")


##export data to showcase in dashboard
saveRDS(SR_map, 
        paste0(output_path, '/SR_map.rds'))

saveRDS(SR_ts, 
       paste0(output_path, '/SR_ts.rds'))


#saveRDS(HSh_map, 
#        paste0(output_path, '/HSh_map.rds'))

#saveRDS(HSh_ts, 
#       paste0(output_path, '/HSh_ts.rds'))

saveRDS(HSi_map, 
        paste0(output_path, '/HSi_map.rds'))

saveRDS(HSi_ts, 
        paste0(output_path, '/HSi_ts.rds'))

```

#8.6. Cumulative Species Richness

```{r}
'This indicator tracks the total number of unique species observed from the beginning of a specified time period up to a given year. It provides an estimation of how many new species are still being recorded over time within a region, helping to evaluate sampling effort and assess the overall recorded biodiversity over the study duration. This is an inherently temporal indicator, presented as a time series (indicator_ts).'

gc()

CR_ts <- cum_richness_ts(Angiosperm_data_cube, 
                         cell_size="grid")

CR_ts_plot <- plot(CR_ts)

##export data to showcase in dashboard

saveRDS(CR_ts, 
        paste0(output_path, '/CR_ts.rds'))

```

#8.7. Mean Year of Occurrence (Newness)

```{r}
'This variable calculates the average year of occurrence for all records within a given spatial unit (e.g., grid cell, for indicator_map) or temporal unit (e.g., year, for indicator_ts), providing an estimation of the relative recency of observations. Maps can highlight areas with more recent average records, while time series show the average observation date over cumulative data.'

#bootstrapping ok

gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
NN_map <- newness_map(Angiosperm_data_cube,
                            cell_size="grid")

#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
NN_ts <- newness_ts(Angiosperm_data_cube, 
                          cell_size="grid")

##export data to showcase in dashboard
saveRDS(NN_map, 
        paste0(output_path, '/NN_map.rds'))

saveRDS(NN_ts, 
        paste0(output_path, '/NN_ts.rds'))

```

#8.8. Occurrence Density

```{r}
'Occurrence Density measures the spatial concentration of records by calculating the total number of occurrences per square kilometre. This allows for more meaningful comparisons between spatial units of different sizes. It can be displayed as a map (indicator_map) illustrating areas with higher concentrations of records, or as a time series (indicator_ts) showing changes in the rate of recording over time for the study area.'

#bootstrapping ok

gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
OD_map <- occ_density_map(Angiosperm_data_cube,
                            cell_size="grid")
OD_map_plot <- plot(OD_map)

#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
OD_ts <- occ_density_ts(Angiosperm_data_cube,
                        cell_size="grid")
OD_ts_plot <- plot(OD_ts)

##export data to showcase in dashboard
saveRDS(OD_map, 
        paste0(output_path, '/OD_map.rds'))

saveRDS(OD_ts, 
        paste0(output_path, '/OD_ts.rds'))
```

#8.9. Species Occurrences

```{r}
'This variable provides the number of occurrences for individual species, focusing on their observed frequency and distribution. It can serve as a proxy for relative abundance for species with similar detectability. It visualizes the observed geographical distribution of a specific species on a map (indicator_map) or tracks its annual record counts over time (indicator_ts).'

#bootstrapping ok

#gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
#SO_map <- spec_occ_map(Angiosperm_data_cube,
 #                           cell_size="grid")
#SO_map_plot <- plot(SO_map)

#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
#SO_ts <- spec_occ_ts(Angiosperm_data_cube, 
#                     cell_size="grid")

#SO_ts_plot <- plot(SO_ts)

##export data to showcase in dashboard
#saveRDS(SO_map, 
#        paste0(output_path, '/SO_map.rds'))

#saveRDS(SO_ts, 
#        paste0(output_path, '/SO_ts.rds'))


```

#8.10. Species Range

```{r}
'Species Range refers to the observed geographical extent of individual species, depicted by the grid cells they occupy. This indicator primarily visualizes the distribution of specific species based on occurrence data, showing occupied cells on a map (indicator_map). Temporally (indicator_ts), it tracks how the number of occupied grid cells for a species changes over time, indicating range expansion, contraction, or stability.'

#bootstrapping ok

#gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
#SRa_map <- spec_range_map(Angiosperm_data_cube,
#                           cell_size="grid")
#SRa_map_plot <- plot(SRa_map)

#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
#SRa_ts <- spec_range_ts(Angiosperm_data_cube, 
#                        cell_size="grid")

#SRa_ts_plot <- plot(SRa_ts)

##export data to showcase in dashboard
#saveRDS(SRa_map, 
#        paste0(output_path, '/SRa_map.rds'))

#saveRDS(SRa_ts, 
#        paste0(output_path, '/SRa_ts.rds'))


```

#8.11. Taxonomic Distinctness

```{r}

'Taxonomic Distinctness measures the average taxonomic distance between any two randomly chosen species in a sample, incorporating their phylogenetic or taxonomic relatedness. A higher value indicates a broader representation of evolutionary history within the community. It can highlight areas on a map (indicator_map) or show trends over time (indicator_ts) where species are, on average, more distantly related.'

#gc()

#TODO: lukt niet om schaal aan te passen naar 1 km
#TD_map <- tax_distinct_map(Angiosperm_data_cube,
#cell_size="grid")
#TD_map_plot <- plot(TD_map)

#TODO;check dat NA buiten Vlaanderen resultaat niet beïnvloedt?
#TD_ts <- tax_distinct_ts(Angiosperm_data_cube,
#                        cell_size="grid")

#TD_ts_plot <- plot(TD_ts)

##export data to showcase in dashboard
#saveRDS(TD_map, 
#        paste0(output_path, '/TD_map.rds'))

#saveRDS(TD_ts, 
#        paste0(output_path, '/TD_ts.rds'))

```

#8.12. Occupancy turnover

```{r}
'Occupancy turnover measures the change in species composition over time, reflecting the rate at which species appear or disappear from a given area. It provides insights into the dynamic nature of ecological communities, highlighting shifts in species distributions and potential environmental changes. High turnover rates may indicate rapid community restructuring, potentially driven by factors such as habitat alteration, climate change, or invasive species. Analyzing occupancy turnover can be crucial for understanding ecosystem stability, identifying areas of conservation concern, and assessing the effectiveness of management strategies.'


gc()

OT_ts <- occ_turnover_ts(Angiosperm_data_cube,
                         cell_size="grid")

OT_ts_plot <- plot(OT_ts)

##export data to showcase in dashboard
saveRDS(OT_ts, 
        paste0(output_path, '/OT_ts.rds'))


```
#8.13 completeness

```{r}
'This function calculates the completeness of biodiversity data over a gridded map or as a time series using the Sample Coverage metric from the iNEXT package.'


gc()

ComP_ts <-completeness_ts(Angiosperm_data_cube,
                         cell_size="grid",
                         gridded_average=TRUE)

ComP_map <-completeness_map(Angiosperm_data_cube,
                            spherical_geometry = FALSE,
                         cell_size="grid")


##export data to showcase in dashboard
saveRDS(ComP_ts,
        paste0(output_path, '/ComP_ts.rds'))

##export data to showcase in dashboard
saveRDS(ComP_map,
        paste0(output_path, '/ComP_map.rds'))


```

#8.14 spec_richness_density

```{r}
'This function calculates the density of unique species over a gridded map or as a time series (see Details for more information).'


gc()

SRD_ts <-spec_richness_density_ts(Angiosperm_data_cube,
                         cell_size="grid")
SRD_map <-spec_richness_density_map(Angiosperm_data_cube,
                        spherical_geometry = FALSE,
                         cell_size="grid")


##export data to showcase in dashboard
saveRDS(SRD_ts,
        paste0(output_path, '/SRD_ts.rds'))

##export data to showcase in dashboard
saveRDS(SRD_map,
        paste0(output_path, '/SRD_map.rds'))


```
