---
title: "Calculate General Indicators with b3gbi and Florabank database on GBIF"
output: html_document
---

#install packages

```{r}
library(b3gbi)
library(rgbif)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
#more info about download here (focussing on dataset of Florabank): https://www.gbif.org/occurrence/download/0063040-260226173443078
download_key <- "0063040-260226173443078"
#download data
dl <- occ_download_get(download_key, overwrite = TRUE)
#import data
dat <- occ_download_import(dl)


#only data before 2023 is complete in GBIF for Florabank
data <- dat%>%filter(year<2023)
```





```{r}
#only maintain cells which are well surveyed according to methodology of Florabank (per year and per 1km gridcell more than 100 unique species observed)
data_filtered <- data %>%
  group_by(eeacellcode, year) %>%            # groep per cel en jaar
  filter(n_distinct(species) > 100) %>%      # enkel groepen met >100 unieke soorten
  ungroup() 


```


##Case study: Cornflower — this species has declined in occupancy due to the use of herbicides, but has recently become more abundant again thanks to its inclusion in seed mixes distributed by the agriculture and fisheries sector as a countermeasure.
```{r}
#specieskey van Korenbloem is 3089561 -> https://www.gbif.org/species/3089561

species_key <- 3089561

# Per jaar berekenen percentage van bezette hokken bezet door de soort en zet waarde naar 0 indien de soort niet aanwezig (dit is gerechtvaardigd aangezien goed onderzochte hoeken)
presence_per_year <- data_filtered %>%
  group_by(year, eeacellcode) %>%
  summarise(present = any(specieskey == species_key), .groups = "drop") %>%
  group_by(year) %>%
  summarise(
    n_cells = n(),
    n_present = sum(present),
    perc_present = mean(present) * 100
  ) %>%
  complete(
    year = full_seq(year, 1),   # vult ontbrekende jaren op
    fill = list(
      n_cells = 0,
      n_present = 0,
      perc_present = 0
    )
  )

grafiek <- ggplot(presence_per_year, aes(x = year, y = perc_present)) +
  geom_point(color = "darkred", size = 3) +
  geom_smooth(method = "loess", color = "steelblue", fill = "lightblue", se = TRUE) +
  scale_y_continuous(limits = c(0, 100)) +   # 🔥 dit is wat je zoekt
  labs(
    title = "Relative occupancy of well serveyed cells",
    x = "Year",
    y = "Percentage (%)"
  ) +
  theme_minimal(base_size = 14)
```



```{r}
#test with b3gbi of absolute occupancy
korenbloem <- data_filtered%>% filter(specieskey== 3089561)

korenbloem_data_cube <- process_cube(korenbloem)

#trend in absolute occupancy over de tijd van korenbloem
plot_ts(spec_range_ts(korenbloem_data_cube),
        title = 'Absolute occupancy of well surveyed cells')

```





##Case study: Bee orchid — this species is becoming more abundant due to increasingly suitable climatic conditions driven by climate change.
```{r}
#specieskey bee orchid is 2792998-> https://www.gbif.org/species/3089561
species_key <- 2792998
presence_per_year <- data_filtered %>%
  group_by(year, eeacellcode) %>%
  summarise(present = any(specieskey == species_key), .groups = "drop") %>%
  group_by(year) %>%
  summarise(
    n_cells = n(),
    n_present = sum(present),
    perc_present = mean(present) * 100
  ) %>%
  complete(
    year = full_seq(year, 1),   # vult ontbrekende jaren op
    fill = list(
      n_cells = 0,
      n_present = 0,
      perc_present = 0
    )
  )

b <- ggplot(presence_per_year, aes(x = year, y = perc_present)) +
  geom_point(color = "darkred", size = 3) +
  geom_smooth(method = "loess", color = "steelblue", fill = "lightblue", se = TRUE) +
  scale_y_continuous(limits = c(0, 5)) +   # 🔥 dit is wat je zoekt
  labs(
    title = "Relative occupancy of well serveyed cells",
    x = "Year",
    y = "Percentage (%)"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
#test with b3gbi of absolute occupancy
bijenorchis <- data_filtered%>% filter(specieskey== 2792998)

bijenorchis_data_cube <- process_cube(bijenorchis)

#trend in absolute occupancy over de tijd van korenbloem
plot_ts(spec_range_ts(bijenorchis_data_cube),
        title = 'Absolute occupancy of well surveyed cells')

```

