---
title: Phylogenetic diversity 5km
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

::: {.callout-tip collapse="true" appearance="minimal" icon="false"}
## Phylogenetic diversity

Phylogenetic diversity (PD) is a measure of biodiversity which takes evolution into account. It is calculated as the sum of the lengths of the phylogenetic tree branches representing the minimum tree-spanning path among a group of species. Phylogenetic diversity can be used in conservation planning to maximise a variety of features, meaning we do not aim to conserve specific features, but rather want to boost a diverse range of features. Conserving a variety of features could be particularly useful in light of the changing environmental conditions, as we can only guess which features will be important in the future.

```{r, echo = FALSE, warning = FALSE, message=FALSE}
library(sf)
library(ggplot2)

pd_map <- read_sf("./pdIndicator/pd_Vl_5km.geojson")

ggplot(data = pd_map) +
  geom_sf(aes(fill = pd), color = NA) +
  scale_fill_viridis_c(na.value = NA) +   # voor continue waarden
  theme_minimal()+
  theme(legend.title = element_blank())
```

Loss map 

```{r, echo = FALSE, warning = FALSE, message=FALSE}
library(sf)
library(ggplot2)

pd_map <- read_sf("./pdIndicator/pd_loss_Vl_5km.geojson")

ggplot(data = pd_map) +
  geom_sf(aes(fill = pd_loss), color = NA) +
  scale_fill_viridis_c(na.value = NA) +   # voor continue waarden
  theme_minimal()+
  theme(legend.title = element_blank())
```

:::
